---
title: "Week 8 Solutions: Contrasts and Estimable Functions"
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: show
---

## Conceptual Solutions

### Solution 1: Estimability Proof

**To prove**: The contrast $\alpha_i - \alpha_j$ is estimable in the effects model $y_{ij} = \mu + \alpha_i + e_{ij}$ with $\sum \alpha_i = 0$.

**Proof**:

For a contrast to be estimable, its coefficient vector must be in the row space of $\mathbf{X}$.

Consider the design matrix for the effects model (example with 3 groups, 2 obs each):

$$
\mathbf{X} = \begin{bmatrix}
1 & 1 & 0 & 0 \\
1 & 1 & 0 & 0 \\
1 & 0 & 1 & 0 \\
1 & 0 & 1 & 0 \\
1 & 0 & 0 & 1 \\
1 & 0 & 0 & 1
\end{bmatrix}
$$

For the contrast $\alpha_1 - \alpha_2$, the coefficient vector is $\mathbf{c}' = [0, 1, -1, 0]$.

Take row 1 of $\mathbf{X}$: $[1, 1, 0, 0]$
Take row 3 of $\mathbf{X}$: $[1, 0, 1, 0]$

Their difference is: $[1, 1, 0, 0] - [1, 0, 1, 0] = [0, 1, -1, 0] = \mathbf{c}'$

Since $\mathbf{c}'$ can be expressed as a linear combination of rows of $\mathbf{X}$, it is in the row space of $\mathbf{X}$, and therefore the contrast $\alpha_1 - \alpha_2$ is estimable. ∎

### Solution 2: Orthogonal Contrasts

Orthogonal contrasts partition SS(Treatments) because they represent **independent** sources of variation.

For $g-1$ orthogonal contrasts:
- Each contrast accounts for 1 degree of freedom
- Together they account for $g-1$ degrees of freedom (same as treatments)
- Since they're orthogonal (independent), their SSs add without overlap

Mathematically: If $\mathbf{c}_1'\mathbf{c}_2 = 0$ (orthogonal), then:
$$\text{SS}(\mathbf{c}_1) + \text{SS}(\mathbf{c}_2) = \text{partial sum of SS(Treatments)}$$

With $g-1$ orthogonal contrasts, we capture all treatment variation.

**For unbalanced designs**: Orthogonality defined as $\mathbf{c}_1'(\mathbf{X}'\mathbf{X})^{-}\mathbf{c}_2 = 0$ is harder to achieve, and the simple summation property may not hold exactly.

### Solution 3: Genetic Evaluation Context

**Testable statement**: "Sire A is different from Sire B"
- This is a **contrast**: Sire A effect minus Sire B effect
- Estimable because it's in the row space of $\mathbf{X}$
- Has a unique estimate regardless of constraints

**Potentially non-estimable**: "Sire A has breeding value +10 kg"
- This is an **absolute value** of a single parameter
- In models with multiple sires, individual sire effects may be confounded with the overall mean
- Different constraints (set-to-zero vs. sum-to-zero) give different absolute values
- Only **differences** between sires are uniquely estimable

## Computational Solutions

### Solution 1: Swine Litter Size

**Complete solution with R code**:

```{r}
#| label: swine-solution
#| message: false

# Data
yorkshire <- c(11, 12, 11)
landrace <- c(10, 11, 10)
duroc <- c(9, 10, 9)

# Part (a): Cell means model
y <- c(yorkshire, landrace, duroc)
breed <- factor(rep(c("Yorkshire", "Landrace", "Duroc"), each = 3))
n <- length(y)

X <- model.matrix(~ breed - 1)
b <- solve(t(X) %*% X) %*% t(X) %*% y
cat("Breed means:\n")
print(b)

# Calculate MSE
residuals <- y - X %*% b
SSE <- sum(residuals^2)
MSE <- SSE / (n - ncol(X))
cat("\nMSE =", MSE, "\n")

# Part (b): Yorkshire vs Landrace
c1 <- c(1, -1, 0)
est1 <- sum(c1 * b)
se1 <- sqrt(sum(c1^2) / 3 * MSE)
t1 <- est1 / se1
p1 <- 2 * (1 - pt(abs(t1), 6))
cat("\n(b) Yorkshire - Landrace:\n")
cat("Estimate:", est1, "SE:", se1, "t:", t1, "p:", p1, "\n")

# Part (c): Yorkshire vs average of others
c2 <- c(1, -0.5, -0.5)
est2 <- sum(c2 * b)
se2 <- sqrt(sum(c2^2) / 3 * MSE)
t2 <- est2 / se2
p2 <- 2 * (1 - pt(abs(t2), 6))
cat("\n(c) Yorkshire - (Landrace+Duroc)/2:\n")
cat("Estimate:", est2, "SE:", se2, "t:", t2, "p:", p2, "\n")

# Part (d): 95% CIs
t_crit <- qt(0.975, 6)
ci1 <- est1 + c(-1, 1) * t_crit * se1
ci2 <- est2 + c(-1, 1) * t_crit * se2
cat("\n(d) 95% Confidence Intervals:\n")
cat("Contrast 1:", ci1, "\n")
cat("Contrast 2:", ci2, "\n")

# Part (e): Check orthogonality
ortho_check <- sum(c1 * c2)
cat("\n(e) Orthogonality: c1' c2 =", ortho_check, "\n")
cat("Contrasts are", ifelse(abs(ortho_check) < 1e-10, "orthogonal", "NOT orthogonal"), "\n")
```

## Applied Solutions

### Solution Template: Broiler Strains

```{r}
#| label: broiler-template
#| eval: false

# Load data (example structure)
# broiler <- read.csv("broiler_strains.csv")

# Define meaningful contrasts:
# c1: Fast-growing commercial vs. Heritage
# c2: Strain A vs. Strain B (within commercial)
# c3: Group comparison
# c4: Specific hypothesis

# For each contrast:
# 1. Compute estimate
# 2. Compute SE
# 3. Compute t-statistic
# 4. Compute p-value
# 5. Interpret biologically

# Example structure:
# test_contrast <- function(c_vec, b, X, MSE) {
#   estimate <- sum(c_vec * b)
#   se <- sqrt(sum(c_vec^2) / n_per_group * MSE)
#   t_stat <- estimate / se
#   p_val <- 2 * (1 - pt(abs(t_stat), df_error))
#   return(c(est = estimate, se = se, t = t_stat, p = p_val))
# }
```

## Notes

These solutions provide frameworks and examples. Students should work through the logic themselves before consulting these answers.

---

**Main Chapter**: [Week 8: Contrasts](Week08_Contrasts.qmd)

**Exercises**: [Week 8 Exercises](Week08_Exercises.qmd)
